---
title: "Billboard Hot 100 & Spotify Audio Analysis"
author: "Simona 
format: html
    embed-resources: true
---

Loading necessary libraries for data manipulation and date handling.
```{r}
library(readr)
library(tidyverse)
library(lubridate)
library(stringr)
```

Loading and inspecting data.
```{r}
billboard <- read_csv("C:/Users/Simona Cholakova/Downloads/archive/charts.csv")
spotify <- read_csv("C:/Users/Simona Cholakova/Downloads/archive (1)/dataset.csv")

glimpse(billboard)
glimpse(spotify)

head(billboard)
head(spotify)
```

Cleaning the data:
```{r}
billboard_clean <- billboard |>
    mutate(
        song_clean = str_to_lower(str_trim(song)),
        artist_clean = str_to_lower(str_trim(artist))
    )

spotify_clean <- spotify |>
    mutate(
        song_clean = str_to_lower(str_trim(track_name)),
        artist_clean = str_to_lower(str_trim(artists))
    )
```

Joining the tables:
```{r}
combined_data <- billboard_clean |>
    inner_join(spotify_clean, by = c("song_clean", "artist_clean"), relationship = "many-to-many") |>
    distinct(song_clean, artist_clean, .keep_all = TRUE)

nrow(combined_data)
```

Q: For each decade, what did the average popular song sound like?

```{r}
#Extract decade from date

getDecade <- function(x){
  year = year(as.Date(x))
  decade = floor((year/10)) * 10
  decade
}
```

```{r}
#Calculate average audio features for each decade

decade_summary <- combined_data |>
    mutate(decade = getDecade(date)) |>
        summarise(
            avg_danceability = mean(danceability, na.rm = TRUE),
            avg_energy = mean(energy, na.rm = TRUE),
            avg_valence = mean(valence, na.rm = TRUE),
            avg_tempo = mean(tempo, na.rm = TRUE),
            avg_loudness = mean(loudness, na.rm = TRUE),
            avg_acousticness = mean(acousticness, na.rm = TRUE),
            avg_instrumentalness = mean(instrumentalness, na.rm = TRUE),
            avg_liveness = mean(liveness, na.rm = TRUE),
            avg_speechiness = mean(speechiness, na.rm = TRUE),
            avg_duration_ms = mean(duration_ms, na.rm = TRUE),
            avg_key = mean(key, na.rm = TRUE),
            avg_mode = mean(mode, na.rm = TRUE),
            .by = decade
        )

decade_summary
```

### Interpretation of the results


Top 20 artists by total weeks on chart:
```{r}
#ranked by total_weeks
top20 <- combined_data |>
  summarise(
    total_weeks = sum(`weeks-on-board`, na.rm = TRUE),
    times_charted = n(),
    best_rank = min(`peak-rank`, na.rm = TRUE),
    .by = artist
  ) |>
  arrange(desc(total_weeks)) |>
  slice_head(n = 20)

top20   
```

### Interpretation of the results


Q: Do songs that reach #1 sound different from songs that don't?
```{r}
number_one_comparison <- combined_data |>
    mutate(was_number_one = if_else(`peak-rank` == 1, "Yes", "No")) |>
    summarise(
        avg_energy = mean(energy, na.rm = TRUE),
        avg_danceability = mean(danceability, na.rm = TRUE),
        avg_valence = mean(valence, na.rm = TRUE),
        avg_tempo = mean(tempo, na.rm = TRUE),
        avg_loudness = mean(loudness, na.rm = TRUE),
        song_count = n(),
        .by = was_number_one
    )

number_one_comparison
```

### Interpretation of the results

Most consistent artists (charted in multiple decades)

```{r}
#ranking by decades active first, then total weeks as tiebreaker

consistent_artists <- combined_data |> 
  mutate(decade = getDecade(date)) |>
  summarise(
    num_decades_appeared = n_distinct(decade), #in how many decades has artists appeared
    avg_danceability = mean(danceability, na.rm = TRUE),
    total_weeks = sum(`weeks-on-board`, na.rm = TRUE), #how many weeks artist was on charts across all their songs
    best_rank = min(`peak-rank`, na.rm = TRUE),
    songs_on_chart = n_distinct(song_clean),
    .by = artist_clean
  ) |>
  filter(num_decades_appeared > 1) |>
  arrange(desc(num_decades_appeared), desc(total_weeks)) |>
  slice_head(n = 20)

consistent_artists

```

### Interpretation of the results

